---
title: "Data Preprocessing"
format: pdf
author: "Arshak Parsa"
date: last-modified
code-block-border-left: "#31BAE9"
---

```{r}
#| warning: false
library(affy)
```


## GEOquery

```{r}
#| eval: false
gset = getGEO("GSE2990", getGPL = FALSE)
saveRDS(gset,"gset.rds")
```

```{r}
gset = readRDS("gset.rds")
mat = exprs(gset[[1]])
# remove Affymetrix control probes from the data
mat = mat[-grep("^AFFX", rownames(mat)), ] 
dim(mat)
dim(pData(gset[[1]]))
```


```{r}
#| eval: false
write.csv(pData(gset[[1]]),"pData.csv")
```

## Another way

I also tried another way (Since I'm new to these things, I don't know what I'm doing).

First extract the `GSE2990_RAW.tar` file and then run the following command in linux to unzip all the gz files

```{bash}
#| eval: false
for f in *.gz; do gunzip $f; done
```

```{r}
#setwd if needed
cell_file = ReadAffy()
mat2 = exprs(cell_file)
dim(mat2)
```

## RMA Normalization

```{r}
#| eval: false
# If affy::rma doesn't work, run the following command 
#(Don't forget to restart R session)
# BiocManager::install("preprocessCore", configure.args = c(preprocessCore = "--disable-threading"), force= TRUE, update=TRUE, type = "source")
write.csv(exprs(affy::rma(cell_file)),"mat_rma.csv",
          row.names = T,col.names = T)
```

```{r}
mat_rma = read.csv("mat_rma.csv",header = T)
rownames(mat_rma)=mat_rma[,1]
mat_rma = mat_rma[,-1]
```

## RNA degradation

```{r}
deg = AffyRNAdeg(cell_file)

plotAffyRNAdeg(deg)
```

I'm not a biologist, but I guess this plot looks fine!

```{r}
#boxplot(cell_file) # Looks weird so I do not plot it
# This one doesn't look good but I don't want to remove any data
hist(cell_file) 

```


## Is `mat_rma` equal to `mat`?

Remove Affymetrix control probes from the data

```{r}
mat_rma = mat_rma[-grep("^AFFX", rownames(mat_rma)), ]
dim(mat_rma)
# check equality of mat and mat_rma
prod(toupper(sub('.{4}$', '', colnames(mat_rma)))==colnames(mat))
prod(rownames(mat)==rownames(mat_rma))
# rownames have different orders
mat_rma = mat_rma[rownames(mat),]
prod(rownames(mat)==rownames(mat_rma))
max(abs(mat-mat_rma))
```

This means GEOquery normalized data is different from `affy::rma`.

Let's try mas5

```{r}
#| eval: false
mat.mas5 = mas5(cell_file)
# Took a LOT OF TIME!
mat.mas5 = mat.mas5[-grep("^AFFX", rownames(mat.mas5)), ]
write.csv(log2(exprs(mat.mas5)),"mat.mas5.csv",
          row.names = T,col.names = T)
# Just DON'T TRY THIS!
```

## Is `GSE2990_suppl_info.txt` valid?

Let's use this file as our metadata

```{r}
metadata = read.delim("GSE2990_suppl_info.txt")
metadata_pData = pData(gset[[1]])
```

```{r}
GotNewInfo = function(x,y){
  # first element indicates that x and y are identical 
  # second element is the number of new datas that are missing in x
  res = c(all(na.omit(x==y))
           ,sum(!(is.na(y)[which(is.na(x)==TRUE)]))
           ,sum(is.na(x))
           ,sum(is.na(y))
           )
  names(res) = c("is.same","new.info","#naX","#naY")
  return(res)
}
GotNewInfo(c(1,2,NA,4,NA,NA),c(1,2,NA,4,5,7))
```

```{r}
# check GEO accession 🗸
all(metadata$geo_accn==metadata_pData$geo_accession)

# check tamoxifen 🗸
all(metadata_pData$description[which(metadata$treatment=="tamoxifen")]==
"Tumor samples from Tamoxifen-only treated primary breast cancer patients")

# check grade 
table(metadata$grade,useNA="always")
table(metadata_pData$`grade:ch1`)
# How is this possible?
# I guess we shouldn't include grade
all(metadata$grade==as.numeric(metadata_pData$`grade:ch1`))

# I don't know anything about node 

# check size 🗸
# metadata_pData has more missing data! Weird!
GotNewInfo(metadata$size,metadata_pData$`size:ch1`)

#check age 🗸
GotNewInfo(metadata$age,as.numeric(metadata_pData$`age:ch1`))

#check er 🗸
GotNewInfo(metadata$er,as.numeric(metadata_pData$`er:ch1`))

#check event.rfs 🗸
GotNewInfo(metadata$event.rfs,as.numeric(metadata_pData$`distant rfs:ch1`))

#check time.rfs 🗸
all(na.omit(metadata$time.rfs-
        as.numeric(metadata_pData$`time rfs:ch1`)/365)<(10^-6))
# so time.rfs is in years

# And I still don't know anything about time.dmfs
```


```{r}
se = SummarizedExperiment::SummarizedExperiment(
    assays = list(expression = mat),
    colData = metadata[,-5]
)

```

```{r}
#| eval: false
saveRDS(se,"cancer189.rds")
```
